---
title: "Symbiont_structure_plot"
author: "Hugo DENIS"
date: "2025-06-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Script to plot the results of symbiont communities variation based on genome wide differentiation (kmer) and ITS2
#Figure 4

#Load packages
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(ggspatial)
library(scatterpie)
library(ggnewscale)
library(ggOceanMaps)
library(vegan)
library(ggrepel)
```

#Load geographic data
```{r}
path="C:/Users/Hugo/Documents/Data/Symbiont_structure_data/"
figpath="C:/Users/Hugo/Documents/Figures/Genomics/Manuscripts_figures/"

#Load sampling site coordinates
site_coordinates=xlsx::read.xlsx(paste0(path,"Sampling_site_coordinates.xlsx"),sheetName = "Feuil1",check.names=F,rowIndex=1:30,colIndex = 1:7,colClasses = c(rep("character",2),rep("numeric",5)))

#Load Millenium project coral reef shapefiles
unep_reef_uncrop<-sf::st_read(paste0(path,"/14_001_WCMC008_CoralReefs2018_v4_1/01_Data/WCMC008_CoralReef2018_Py_v4_1.shp"))

#Load Western Pacific countries border 
australia_border<-raster::getData("GADM",country="AUS",level=0)
ncl_border<-raster::getData("GADM",country="NCL",level=0)
van_border<-geodata::gadm(country="VUT",level=0,path=path)
van_border=sf::st_as_sf(van_border)

#Load World countries border
world=rnaturalearth::ne_countries(returnclass = "sf")["geometry"]
```

#Load metadata & environmental data
```{r}
meta_colony=read.table(paste0(path,"Colony_metadata_Aspat_GBR_NC.txt"),header=T,sep="\t")
```

#Load symbiont genomic data 
```{r}
#Load symbiont ITS2 annotation
cluster_df=read.table(paste0(path,"D2s_50M_dist_matrix_aspat_clean_cluster_metadata_outliers_1.5IQR_per_region.txt"),check.names=F,sep="\t",header=T)

cluster_df$ITS2[cluster_df$ITS2=="C50b" & cluster_df$Genomic.cluster=="GBR_South"]="C50b_SGBR"
cluster_df$ITS2[cluster_df$ITS2=="C50b" & cluster_df$Genomic.cluster=="NC"]="C50b_NC"

#Load result of dbRDA
dbRDA=readRDS(paste0(path,"Symbiont_full_dbRDA.rds"))
```


#Plot Symbionts populations on MAP 
#Make map with distribution of symbionts clusters as per the host 
```{r}
world_map=ggplotGrob(ggplot(data = world) +
    geom_sf(fill="white",colour="black")+
  geom_rect(aes(xmin = 2564417, xmax = 5174417, ymin = 1564417, ymax = 3064417), color = "#092347", fill = NA,size=1) +
    coord_sf(crs = "+proj=laea +lat_0=-10 +lon_0=160 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs") +theme_void())

site_map=basemap(limits = c(144,169, -26,-13),
        bathymetry = TRUE,land.col=NA,land.border.col=NA,
        glaciers = FALSE,legend.position="bottom",bathy.style = "rcb",bathy.alpha=0.5)  + geom_polygon(data=ncl_border, aes(long,lat,group=group),fill="grey",colour="grey40",size=0.3) + geom_polygon(data=australia_border, aes(long,lat,group=group),fill="grey",colour="grey40",size=0.3)+ geom_sf(data=van_border ,fill="grey",colour="grey40",size=0.3)+ geom_sf(data = unep_reef_uncrop['LAYER_NAME'],fill="grey",color="grey40",xlim=c(144,167),ylim=c(-25,-12),size=0.3)+ theme(legend.title = element_text(size=17,color="black",margin=margin(t = 0, r = 7, b = 35, l = 0)),legend.text = element_text(size=17,color="black"),legend.background = element_rect(fill = "transparent",colour = "transparent"),legend.key.height = unit(4,"mm"),legend.key.width = unit(20,"mm"),legend.position = "none",legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,0,0),axis.text.x = element_text(size=30,color="#092347",vjust = -0.7,hjust=-0.2,
                                 margin = margin( t = -45)),axis.text.y=element_text(size=30,color="#092347",hjust = -0.9,vjust=-0.3,margin=margin(l=0,r=-80,b=0,t=0)),plot.background= element_rect(fill="white", colour="white"),panel.background = element_rect(fill="transparent", colour="#092347",linewidth=1.1),axis.ticks =element_blank())+ labs(x="",y="")+
  scale_y_continuous(labels = ~ paste0(.x, "\u00B0"))+
  scale_x_continuous(labels = ~ paste0(.x, "\u00B0"))+
  ggspatial::annotation_scale(
    location = "tr",
    bar_cols = c("#092347", "white"),
    text_family = "ArcherPro Book",line_width = 1.5,text_cex=4,pad_y = unit(0.4, "in")
  ) + ggspatial::annotation_north_arrow(location = "tr", which_north = "true", pad_x = unit(0.4, "in"), pad_y = unit(0.8, "in"),height=unit(4,"cm"),width=unit(4,"cm"), style = north_arrow_fancy_orienteering(fill = c("#092347", "white"),text_size=50)) + coord_sf(xlim=c(144,169),ylim=c(-26,-13),expand=F,crs="+proj=longlat +datum=WGS84 +no_defs")+ geom_point(data=site_coordinates,aes(x=Lon,y=Lat),size=5,shape=21, color="#092347",fill="#092347") + guides(fill="none",color="none")+ theme(panel.grid = element_blank())

#Remove samples that could not be assigned an ITS2 type
cluster_df=cluster_df %>% subset(!is.na(Cluster))

#Combine with metadata and compute average per site 
cluster_df$Lon=as.numeric(cluster_df$Lon)
cluster_df$Lat=as.numeric(cluster_df$Lat)
cluster_df_site=cluster_df %>% group_by(locationID) %>% mutate(Lon=mean(Lon,na.rm=T),Lat=mean(Lat,na.rm=T)) %>% ungroup() %>%  group_by(locationID,ITS2) %>% dplyr::summarize(Occurence=n(),Lat=mean(Lat,na.rm=T),Lon=mean(Lon,na.rm=T)) %>% pivot_wider(names_from = "ITS2",values_from = "Occurence", values_fill = list(Occurence = 0))

#Add sites coordinates and adjusted positions to plot the pies on the map
cluster_df_site=cbind(cluster_df_site,data.frame(x.adj=c(-1,1,1,-1,1,2,1,-0.5,1,2,1,-1.5,-1,-1,-1,1.5,2,-1,-0.6,1,2,1,-1,1,2,1,1,-1,1),y.adj=c(-0.2,-0.5,1,1,0.5,-0.5,-1,0,0,0.5,1,-1,0,0.5,0,0.5,-1,-0.4,-0.2,0,0.5,-0.3,0,0.5,1,0.5,0,-0.5,0)))
cluster_df_site=cluster_df_site %>% mutate(Lon.adj=Lon+x.adj,Lat.adj=Lat+y.adj)


cluster_colors=c("#6e350f","#D38B46","#4ec49b","#7683fe","#e76f51")
names(cluster_colors)=c("C3k","C3k/C3bo","C50b_SGBR","C50b_NC","C50c")

cluster_df_site$Tot=rowSums(cluster_df_site[,4:8])
cluster_df_site[,4:8]=cluster_df_site[,4:8]/cluster_df_site$Tot

site_map_pie=site_map + geom_segment(data=cluster_df_site,aes(x=Lon,xend=Lon.adj,y=Lat,yend=Lat.adj),size=0.7,colour="#092347") + new_scale_fill()  + geom_scatterpie(data=cluster_df_site,aes(x=Lon.adj,y=Lat.adj),cols=c("C3k/C3bo","C50c","C3k","C50b_SGBR","C50b_NC"),pie_scale = 0.9,colour=NA) + scale_fill_manual(values=cluster_colors) + coord_sf(xlim=c(144,169),ylim=c(-26,-13),expand=F,crs="+proj=longlat +datum=WGS84 +no_defs") + theme(legend.key.height = unit(10,"mm"),legend.key.width = unit(10,"mm"),legend.position = "none") + labs(fill="ITS2 type")

full_site_map<-site_map_pie+ annotation_custom(grob=world_map,xmin=145,xmax=149.5,ymin=-Inf,ymax=-20)
```

#Plot RDA results
```{r}
#Scaling 1
site_scores_scaling1 <- as.data.frame(vegan::scores(dbRDA, display = "sites", scaling = 1,choices=c(1,2,3,4,5)))
species_scores_scaling1 <- as.data.frame(vegan::scores(dbRDA, display = "species", scaling = 1,choices=c(1,2,3,4,5)))
env_scores_scaling1 <- as.data.frame(vegan::scores(dbRDA, display = "bp", scaling = 1,choices=c(1,2,3,4,5)))

# Add labels for sites, species, and environmental variables
site_scores_scaling1$Site <- rownames(site_scores_scaling1)
species_scores_scaling1$Species <- rownames(species_scores_scaling1)
env_scores_scaling1$Variable <- rownames(env_scores_scaling1)

#Edit variable names 
variable_labeller=c("PC1"="Host PC1","PC2"="Host PC2","PC3"="Host PC3","Kd490_median"="Turbidity Median","CF_median"="Cloud Fraction\nMedian","DHW_collection"="DHW\ncollection","SST_AR_1km"="SST AR","SSTmean_change"="SST change","DHW_freq_sup4_5km"="DHW>4 Freq","DHW_max_5km"="DHW max","Km_to_coastline"="Shore\ndistance","MMM_5km"="MMM","DHW_freq_sup8"="DHW>8 Freq")

#Add percentage of variation explain by each axis
eigenvalues <- dbRDA$CCA$eig
# Calculate the percentage of variance explained by each axis
variance_explained <- eigenvalues / sum(eigenvalues) * 100

#Add cluster information
site_scores_scaling1=site_scores_scaling1 %>% merge(cluster_df %>% dplyr::select(wgsID_short,ITS2),by.x="Site",by.y="wgsID_short") %>% merge(meta_colony %>% dplyr::select(wgsID_short,Geographic_cluster),by.x="Site",by.y="wgsID_short") %>% subset(!is.na(ITS2))

#Retain only environmental variables that were found significant through forward selection
env_scores_scaling1=env_scores_scaling1 %>% subset(Variable %in% c("MMM_5km","DHW_freq_sup4_5km","SST_AR_1km"))


cluster_colors=c("#6E3610","#D38B46","#4EC49B","#00724E")
names(cluster_colors)=c("C3k","C3k/C3bo","C50b","C50c")

cluster_colors=c("#6e350f","#D38B46","#4ec49b","#7683fe","#e76f51")
names(cluster_colors)=c("C3k","C3k/C3bo","C50b_SGBR","C50b_NC","C50c")


#Adjust size of predictor length
dbRDA_plot=function(var,axis1,axis2,coef,v1,v2,h1,h2){
  Coef=coef
text_size=30

dbRDA1_percent <- round(variance_explained[axis1], 1)
dbRDA2_percent <- round(variance_explained[axis2], 1)

env_scores_scaling1$CAP1scaled=Coef*env_scores_scaling1$CAP1
env_scores_scaling1$CAP2scaled=Coef*env_scores_scaling1$CAP2
env_scores_scaling1$CAP3scaled=Coef*env_scores_scaling1$CAP3
env_scores_scaling1$CAP4scaled=Coef*env_scores_scaling1$CAP4
env_scores_scaling1$CAP5scaled=Coef*env_scores_scaling1$CAP5


#Adjust position of eigenvectors labels 
env_scores_scaling1$vjust[env_scores_scaling1[,paste0("CAP",axis2)]>0]=v1
env_scores_scaling1$vjust[env_scores_scaling1[,paste0("CAP",axis2)]<0]=v2
env_scores_scaling1$hjust[env_scores_scaling1[,paste0("CAP",axis1)]>0]=h1
env_scores_scaling1$hjust[env_scores_scaling1[,paste0("CAP",axis1)]<0]=h2

if(axis1==3){
  env_scores_scaling1$hjust[env_scores_scaling1$Variable=="SST_AR_1km"]=0.9
  env_scores_scaling1$vjust[env_scores_scaling1$Variable=="MMM_5km"]=1.2
}

#Change variable names 
env_scores_scaling1$LabelledVariable <- variable_labeller[env_scores_scaling1$Variable]

# Plot for Scaling 1 (Distance Scaling)
plot_scaling1 <- ggplot() +
  # Plot site scores
  geom_point(data = site_scores_scaling1, aes_string(x = paste0("CAP",axis1), y = paste0("CAP",axis2),color=var,shape="Geographic_cluster"), size = 5,alpha=0.8)  +
  # Plot environmental vectors
  geom_segment(data = env_scores_scaling1, aes_string(x = 0, y = 0, xend = paste0("CAP",axis1,"scaled"), yend = paste0("CAP",axis2,"scaled")), 
               arrow = arrow(length = unit(0.6, "cm")), color = "black", size = 1,alpha=0.8) +
  geom_label(data = env_scores_scaling1, aes_string(x = paste0("CAP",axis1,"scaled"), y = paste0("CAP",axis2,"scaled"), label = "LabelledVariable",vjust="vjust",hjust="hjust",segment.colour=NA), 
            color = "black",size=text_size/3) +
  
  # Labels and theme
  labs(x = paste0("dbRDA",axis1," (",dbRDA1_percent,"%)"), y = paste0("dbRDA",axis2," (",dbRDA2_percent,"%)"), title = "") + theme_classic()+ theme(axis.text.x=element_text(size=text_size,color="black",angle=0,hjust=0.95),axis.text.y=element_text(size=text_size,color="black"),axis.title.x = element_text(size=text_size,color="black"),axis.title.y = element_text(size=text_size,color="black",margin = margin(t = 0, r = 7, b = 0, l = 0)),legend.title = element_text(size=text_size-2,color="black",margin = margin(t = 0, r = 20, b = , l = 0)),legend.text = element_text(size=text_size-5,color="black"),legend.position ="bottom",legend.background = element_rect(fill = "transparent"),panel.background = element_rect(colour = "black", size=1.5),legend.key = element_blank(),axis.ticks.length = unit(0.3, "cm"),axis.ticks = element_line(size = 1))

return(plot_scaling1)
  
}

#Plot RDA1 & RDA2
dbRDA_plot_RDA12=dbRDA_plot("ITS2",1,2,3,-0.2,1.2,0.4,0.3)+ggplot2::scale_color_manual(values=cluster_colors,name="Major ITS2")+ggplot2::scale_shape_manual(values=c(16, 17, 15, 18),name="Host population",labels=as_labeller(c("CB"="CB","South GBR"="S GBR","Central/North GBR"="CN GBR","NC"="NC")))  + guides(colour = F,shape=F)

dbRDA_plot_RDA23=dbRDA_plot("ITS2",2,3,3,-0.2,1.2,0.5,0.5)+ggplot2::scale_color_manual(values=cluster_colors,name="Major ITS2")  +ggplot2::scale_shape_manual(values=c(16, 17, 15, 18),name="Host population",labels=as_labeller(c("CB"="CB","South GBR"="S GBR","Central/North GBR"="CN GBR","NC"="NC"))) + guides(colour = F,shape=F)

dummy_plot=dbRDA_plot("ITS2",2,3,4.5,-0.2,1.2,0.5,0.5)+ggplot2::scale_color_manual(values=cluster_colors,name="Major\nITS2",labels = c(
  "C3k"="C3k",
  "C3k/C3bo"="C3k/C3bo",
  "C50b_SGBR"="C50b",
  "C50b_NC"="C50b",
  "C50c"="C50c"
)
)  +ggplot2::scale_shape_manual(values=c(16, 17, 15, 18),name="Host\npopulation",labels=as_labeller(c("CB"="CB","South GBR"="S GBR","Central/North GBR"="CN GBR","NC"="NC")))  + theme(legend.key.width =unit(3,"cm"),legend.text=element_text(size=28,color="black",margin = margin(l = -20, unit = "pt")),legend.title=element_text(size=30,color="black",margin = margin(l = 30, unit = "pt")),legend.spacing.x = unit(0, "mm")) +
  guides(shape = guide_legend(override.aes = list(size = 15)),color = guide_legend(override.aes = list(size = 15)))
legend <-ggpubr::get_legend(dummy_plot)

```

#Combine plot together and save 
```{r}
map_dbRDA_combine_plot=cowplot::ggdraw() +draw_plot(full_site_map,x=0.025,y=0.45,width=1,height=0.55) +draw_plot(dbRDA_plot_RDA12,x=0.02,y=0.05,width=0.46,height=0.4) +draw_plot(dbRDA_plot_RDA23,x=0.505,y=0.05,width=0.46,height=0.4) + theme(plot.background = element_rect(fill="white",color="white"))+draw_plot_label(label=c("a","b"),x=c(0.03,0.03),y=c(1,0.45),size=50) + draw_plot(legend,x=0,y=0,width=1,height=0.05)

ggsave(map_dbRDA_combine_plot ,filename=paste0("C:/Users/Hugo/Documents/Figures/Genomics/Manuscripts_figures/","Figure_4_withoutlabels_v3.png"),width=22.5,height=20,dpi=300)
```
